---
title: "Toronto's continuously growth reported crime cases in past years."
subtitle: "An Analysis of Toronto's reported crime from 2014 to 2022"
author: 
  - Ziheng Zhong
thanks: "Code and data are available at: https://github.com/iJustinn/Totonto_Reported_Crimes.git"
date: today
date-format: long
abstract: "This study presents a comprehensive analysis of reported crime data in Toronto, as sourced from Open Data Toronto, to explore the trend of urban criminal patterns in Toronto over the past years. By examining reported incidents from 2014 to 2022, this paper identifies key trends in crime cases across different years and police divisions, highlighting possible key factors that may caused such results. In a word, the paper seeks to provide actionable insights for city planners, law enforcement, and community advocates, aiming to build a safer, more enjoyable urban living environment."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(opendatatoronto)
library(tidyverse)
library(ggplot2)
library(janitor)
library(dplyr)
library(readr)

# read the sum_by_year_data, ready for generating bar chart
sum_by_year_data <- read_csv(here::here("outputs/data/sum_by_year_data.csv"))

# read the sum_by_year_data, ready for generating bar chart
sum_by_division_data <- read_csv(here::here("outputs/data/sum_by_division_data.csv"))
```



# Introduction

Public safety, as one of the most important factors to ensure a high quality life for people, is an important aspect that requires continuous evaluation and enforcement. The city of Toronto, as one of the place with most population in Canada, is not an exception to the challenges posed by crimes. This paper takes a deep dive into the extensive reported crime data provided by Open Data Toronto[@citeODT], covering the period from 2014 to 2022. The analysis is grounded in a methodical approach to identify and interpret crime trends across the past years, also within the city's diverse police divisions, offering a detailed chronicle of criminal activity over selected time span. To be more specific, this research analyzed and visualized how crime spreads out across different divisions and how it changes over the years. By dissecting the data year by year, division by division, this study uncovers patterns that could inform future urban planning and law enforcement strategies. The role of population growth, natural diseases and urban development, are considered as potential catalysts influencing these patterns.

Moreover, the paper aims to a little beyond simply displaying statistics. It endeavors to contextualize the numerical findings within the broader spectrum of Toronto's urban structure. The motivation for this discussion is that changes in crime rates are often associated with changes in other indicators of the city, such as education level[@citeEdu], unemployment rate[@citeUnemp], economic level[@citeEcon], and so on. By doing this, the study aims to offer useful advice that can help city officials and the police create plans to make the city safer and better for all residents. Through these ways, the paper is not just an exposition of data, but a roadmap towards a better living environment.

The subsequent sections of this paper are organized to facilitate a comprehensive understanding of the study and its implications. Following this introduction, @sec-2 outlines the methodology employed in the analysis, ensuring transparency and replicability. @sec-3 presents the findings in detail, charting the trends in reported crimes over time and across Toronto's police divisions. also discusses these findings in other aspects such as: infrastructure construction, structure of the city, etc. Finally, @sec-4 concludes the paper, summarizing the key insights and finding in this research.



# Data {#sec-2}

## Source of Data

## Methods

Some of our data is of 







# Discussion {#sec-3}

## Across years
```{r}
#| label: by_year_chart
#| fig-cap: 2014 to 2022 Toronto Total Reported Crime Cases by Year
#| echo: false

# create the bar chart with y-axis showing total cases in the year and x-axis showing the year number
ggplot(sum_by_year_data, aes(x = report_year, y = total_cases)) +
  scale_x_continuous(breaks = sum_by_year_data$report_year) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(x = "Report Year",
       y = "Total Cases") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



## Across divisions
```{r}
#| label: by_division_chart
#| fig-cap: 2014 to 2022 Toronto Total Reported Crime Cases by Division
#| echo: false

# create the bar chart with y-axis showing total cases in the division and x-axis showing the division number
ggplot(sum_by_division_data, aes(x = division, y = total_cases, fill = division)) +
  #geom_bar(stat = "identity", position = position_dodge()) +
  geom_bar(stat = "identity", fill = "darkgrey") +
  theme_minimal() +
  labs(x = "Police Division",
       y = "Total Cases") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


# Conclusion {#sec-4}

## General



## Other aspects




